Poker Chips, Earthquakes, and Computation

### 2014-10-11 finaldraft.tex
Fixed some errors
Also started working on bringing d tree section in line on notation.

To define the spread function, we fix one coordinate on a board $b$ as the origin $(0,0)$, and we define the \emph{taxicab distance} $d(s)$ of a node $s$ on the board to be $|s_x| + |s_y|$, where $s_x$ is the $x$ coordinate of node $s$, and $s_y$ is the $y$ coordinate of node $s$. Then the spread function is
\[ \psi(b) = \sum_{\text{all nodes }s}b(s)d(s)^2 \]

Thus, the function sums up the chip stack sizes weighted by the taxicab distance to the origin squared. The motivation for this definition is that the spread function gives a measure of how ``spread out" the chips on a board are. Intuitvely, earthquakes move chips that are condensed in a single pile and spreads them out over a greater area. We capture this in the following lemma:

\begin{lemma}
\label{spreadfunction}
Suppose an earthquake hits node $s$ of a board $b$, producing the board $E_sb$. Then
\[ \Psi(E_sb) > \Psi(b) \]
so the earthquake strictly increases the spread of the board.
\end{lemma}
\begin{proof}
The chip stacks of the two boards are the same at all nodes except at $s$ and $n(s)$, the nodes that neighbor $s$. Thus, the contribution to $\Psi$ is the same at all nodes except for those 5.

Let the taxicab distance to $s$ is $d_s$. Then the contribution to the spread function from node $s$ decreases from $b(s)d_s^2$ to $(b(s)-4)d_s^2$.

Out of the 4 nodes in $n(s)$, some have a taxicab distance of $d_s+1$, and the rest are at distance $d_s-1$. In particular, if $s$ is not exactly on the $x$ or $y$ axis, exactly 2 neighboring nodes are at $d_s+1$, and exactly 2 are at distance $d_s-1$. If $s$ is on either the $x$ or $y$ axis, then 3 nodes are at distance $d_s+1$, and 1 neighboring node is at distance $d_s-1$. Since we want to lower bound the increase in the spread function, we focus only on the case where exactly 2 neighboring nodes are at distance $d_s+1$, as it can be shown that this case leads to a strictly smaller increase than the other.

@@

Omitting the algebraic details, we find that summing up the changes in the spread function gives
\[ \Psi(E_sb) - \Psi(b) = 4\]
which completes the proof.
\end{proof}

\begin{corollary}
\label{finitestability}
A board can never return to a previous state via a sequence of earthquakes.
\end{corollary}
\begin{proof}
Suppose for the sake of contradiction that this is possible, and that $b = E_1E_2\cdots E_kb$ for some sequence of earthquakes $E_1,E_2,\dots,E_k$. Lemma~\ref{spreadfunction} shows that $\psi(b) > \psi(E_1E_2\cdots E_kb)$, which is a contradiction.
\end{proof}

@@

We are now ready to prove the theorem that all finite boards will reach a stable state.

\begin{proof}
(of Theorem~\ref{finitestability}) Recall that $b$ is a finite board so $\sum_{i,j} b(i,j) = C$, furthermore this tells us that the locations with a nonzero number of chips are bounded.  Without loss of generality we can say that $b$ is nonzero only in the range $[i_{min}, i_{max}] \times [j_{min}, j_{max}]$. Then by Lemma~\ref{finiteextension}, we have that after any number of earthquakes the resultant board is nonzero only in the range $ R=[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$. If we set 
\begin{equation*}
D= |j_{max}-j_{min} + 2C|*|i_{max} - i_{min} + 2C|
\end{equation*}
then there are at most  $\binom{D+C-1}{D-1}$  boards that have exactly $C$ chips on them and are nonzero only in $R$,. We know that the state of the board cannot cycle and there are only a finite number of board states so eventually the board must stabilize.
\end{proof}

@@

\begin{definition}
We say that an earthquake $E$ is valid if the node it targets has at least $4$ chips on it. 
We say that a sequence of earthquakes $E_1, E_2, \ldots E_n$ to be a valid sequence of earthquakes if each earthquake is valid in the order applied. That is 
\begin{equation*}
E^{m}[b, \{s_k\}](s_{m+1}  \geq 4
\end{equation*}
for all $1\geq m+1 \leq n$.
\end{definition}

@@

Recalling that $n(s)$ refers to the set of neighbors of the square $s$ we note that if an earthquake $E$ is valid then we can write it in terms of this indicator function.
\begin{lemma} 
\label{earthquakeredefine}
Consider a board $b$ if $E$ is a valid earthquake, then 
\begin{equation*}
E[b, s](x) = b(x) + 1[[ x \in n(s)]] - 4[[x =s]]
\end{equation*}
\end{lemma}
\begin{proof}
This follows directly from the definition of an earthquake and a valid earthquake. 
\end{proof}

Now we can show that switching the order of two earthquakes with certain validity conditions does not affect the final board state.
\begin{lemma}
\label{swappinglemma}
Consider a board $b$ and two earthquakes $E_1$ and $E_2$, affecting nodes $s_1$ and $s_2$ respectively. If $E_1, E_2$ is a valid sequence and $E_2$ is a valid earthquake when applied before $E_1$ then $E_2, E_1$ is also a valid sequence of earthquakes and 
\begin{equation*}
E_2[E_1[b, s_1] s_2] = E_1[LE_2[b, s_2], s_1] 
\end{equation*}
\end{lemma}
\begin{proof}
We will first show that $E_2, E_1$ is a valid sequence and then use this to show that the order in which the earthquakes occur does not matter.

We demonstrate this by contradiction. Assume  $E_2, E_1$ is an invalid sequence, that is $E_2[b, s_2](s_1)<4$. Since $E_1$ is a valid earthquake when applied directly to $b$ we know $b(s_1) \geq 4 $. 
Therefore we must have $b(s_1)<8$ and $s_2=s_1$. But then $E_1[b, s_1](s_1)<4$ and so $E_1, E_2$ would not be a valid sequence of earthquakes. This contradicts one our hypothesis and so $E_2, E_1$ is a valid sequence.

Now in order to demonstrate that the order in which the earthquakes are applied does not affect the final board state consider $E_2[ E_1[b, s_1], s_2](x)$ for some $x \in \mathbb{Z} \times \mathbb{Z}$. Applying Lemma ~\ref{earthquakeredefine} we have 
\begin{align*}
E_2[E_1[b, s_1], s_2](x) = E_1[b, s_1])(x)  + [[ x \in n(s_2) ]] - 4[[ x=s_2]] \\
= b(x) + [[ x \in n(s_1) ]] - 4[[x = s_1] + [[ x \in n(s_2) ]] - 4[[ x= s_2)]]  \\
= b(x) + [[ x \in n(s_2) ]] - 4[[x = s_2] + [[ x \in n(s_1) ]] - 4[[ x= s_1)]]\\
= E_2[b, s_2](x) + [[ x \in n(s_1) ]] - 4[[ x = s_1]] \\
= E_1[E_2 [b, s_2], s_1](x) 
\end{align*}
The critical step in this proof is recognizing that because $E_1$ and $E_2$ are valid earthquakes regardless of their ordering the algebra above was valid.
\end{proof}

@@

\begin{definition}
A $d$-tree is a graph composed of a parent node connected by edges to $d-1$ unique children nodes.Those children each have $d-1$ unique children and so on.  
\end{definition}
We note that every node other than the original parent node has $d$ neighbors. In order to simplify our process we would like the graph to be regular and so we make the following definition. 
\begin{definition}
A joined $d$-tree is two $d$-trees connected at their parent nodes by an edge.
\end{definition}

@@

In order to keep our notation compact we once again define a function that outputs the neighbors of a node
\begin{definition} 
We define the function $n: G_d \rightarrow G_d^d$ as 
\begin{equation*}
n(i,j) = \{ (x_1, y_1); (x_1 = i+1 \text{ or } x_1=i-1), \text{ and }(x_1, y_1) \in G_d\}
\end{equation*}
\end{definition}

We now formalize the board of poker chips. Note that this definition only differs from our initial definition in that the input value is a coordinate of a node in a joined $d$-tree rather than a coordinate in the integer lattice.
\begin{definition}
We define a $d$-board to be a map from the set of valid coordinates to the nonnegative integers, that is 
\begin{equation}
b:  G_d \rightarrow \mathbb{Z}_{\geq 0}
\end{equation}
We write $\mathcal{B}(d)$ to refer to the set of all $d$-boards.  
\end{definition}

Recalling that $n(s)$ refers to the neighbors of $s$ we define our earthquake operation in terms of $d$-boards.
\begin{definition}
An earthquake takes as an input an $d$-board and a node and outputs an $d$-board so we write $E: \mathcal{B}(d) \times G_d \rightarrow \mathcal{B}(d)$. If $s$ is the node input and $b(s)>d$ then 
\begin{align*}
E[b, s](s) = b(s)- d \\
E[b, s](t) = b(t) +1 
\end{align*}
for $t \in n(s)$. Otherwise $b(s)<d$ and $E[b,s]=b$ as the earthquake does not move any chips. We use the notation $E^r[b, \{s_k\}]$ to denote the application of $E$ to the $b$ $r$ times, to the sites $s_1, \ldots s_r$ respectively.
\end{definition}

Now that we have formalized the earthquake operator we can introduce a precise definition of stability of a board. 
\begin{definition}
We say that a $d$-board is stable if its values are strictly less than $n$.
\end{definition}
This is equivalent to saying all earthquakes have no effect on the board. We also make a precise definition for a finite board.
\begin{definition} 
We say that a $d$-board is finite if the total number of chips on it is finite. That is a $d$board $b$ is finite if the sum
\begin{equation}
C= \sum_{(i,j) \in G_d} b(i,j) 
\end{equation}
is finite. 
We denote the set of finite $d$-boards as $B(d)_F$.
\end{definition}

@@

We once again define a spread function 
\begin{definition}
 Let $\Psi: \mathcal{B}(d)_F \rightarrow \mathbb{Z}_{\geq 0}$ be the function
\begin{equation}
\Psi(f) = \sum_{(i,j) \in G_d} b(i,j) |i|^2
\end{equation}
Note that this is well defined for trees which contain a finite number of chips. 
\end{definition} 
We can once again think of this spread function as the sum of the number of chips on every node weighted by the taxi-cab distance from the node to the origin.

>>>>

To define the spread function, we fix one coordinate on a board $b$ as the origin $(0,0)$, and we define the \emph{taxicab distance} $d(s)$ of a node $s$ on the board to be $|s_x| + |s_y|$, where $s_x$ is the $x$ coordinate of node $s$, and $s_y$ is the $y$ coordinate of node $s$. Then the spread function is
\begin{equation*}
\Psi(b) = \sum_{\text{all nodes }s}b(s)d(s)^2 
\end{equation*}

Thus, the function sums up the chip stack sizes weighted by the taxicab distance to the origin squared. The motivation for this definition is that the spread function gives a measure of how ``spread out" the chips on a board are. Intuitively, earthquakes takes chips that are condensed in a single pile and spreads them out over a greater area. We capture this in the following lemma:

\begin{lemma}
\label{spreadfunction}
Suppose an earthquake hits node $s$ of a finite board $b$, producing the board $E_sb$. Then
\begin{equation*}
\Psi(E_sb) > \Psi(b) 
\end{equation*}
so the earthquake strictly increases the spread of the board.
\end{lemma}
\begin{proof}
The chip stacks of the two boards are the same at all nodes except at $s$ and $n(s)$, the nodes that neighbor $s$. Thus, the contribution to $\Psi$ is the same at all nodes except for those 5.

Let the taxicab distance to $s$ be $d_s$. Then the contribution to the spread function from node $s$ decreases from $b(s)d_s^2$ to $(b(s)-4)d_s^2$.

Out of the 4 nodes in $n(s)$, some have a taxicab distance of $d_s+1$, and the rest are at distance $d_s-1$. In particular, if $s$ is not exactly on the $x$ or $y$ axis, exactly 2 neighboring nodes are at $d_s+1$, and exactly 2 are at distance $d_s-1$. If $s$ is on either the $x$ or $y$ axis, then 3 nodes are at distance $d_s+1$, and 1 neighboring node is at distance $d_s-1$. Since we only need a lower bound for the increase in the spread function, we focus only on the case where exactly 2 neighboring nodes are at distance $d_s+1$, as it can be shown that this case leads to a strictly smaller increase than the other.

@@

Omitting the algebraic details, we find that summing up the changes in the spread function gives
\begin{equation*}
 \Psi(E_sb) - \Psi(b) = 4
 \end{equation*}
which completes the proof.
\end{proof}

\begin{corollary}
A finite board can never return to a previous state via a sequence of earthquakes.
\end{corollary}
\begin{proof}
Suppose for the sake of contradiction that this is possible, and that $b = E_1E_2\cdots E_kb$ for some sequence of earthquakes $E_1,E_2,\dots,E_k$. Lemma~\ref{spreadfunction} shows that $\psi(b) < \psi(E_1E_2\cdots E_kb)$, which is a contradiction.
\end{proof}

@@

We are now ready to show that all finite boards will reach a stable state.

\begin{theorem} If $b$ is a finite board then there exists some sequence $\{s_1, \ldots, s_k\}$ such that after applying an earthquake to each of these locations on $b$ the resultant board is stable. 
\end{theorem}
\begin{proof}
 Since $b$ is a finite board so $\sum_{i,j} b(i,j) = C$, furthermore this tells us that the locations with a nonzero number of chips are bounded.  Without loss of generality we can say that $b$ is nonzero only in the range $[i_{min}, i_{max}] \times [j_{min}, j_{max}]$. Then by Lemma~\ref{finiteextension}, we have that after any number of earthquakes the resultant board is nonzero only in the range $ R=[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$. If we set 
\begin{equation*}
D= |j_{max}-j_{min} + 2C|*|i_{max} - i_{min} + 2C|
\end{equation*}
then there are at most  $\binom{D+C-1}{D-1}$  boards that have exactly $C$ chips on them and are nonzero only in $R$,. We know that the state of the board cannot cycle and there are only a finite number of board states so eventually the board must stabilize.
\end{proof}

@@

\begin{definition}
We say that an earthquake $E$ is valid if the node it targets has at least $4$ chips on it. 
We say that a sequence of earthquakes $E_1, E_2, \ldots E_n$ to be a valid sequence of earthquakes if each earthquake is valid when applied one after another. 
\end{definition}

@@

Recalling that $n(s)$ refers to the set of neighbors of the square $s$ we note that if an earthquake $E$ is valid then we can write it in terms of this indicator function.
\begin{lemma} 
\label{earthquakeredefine}
Consider a board $b$ if $E$ is a valid earthquake, then 
\begin{equation*}
E_s b(x) = b(x) + 1[[ x \in n(s)]] - 4[[x =s]]
\end{equation*}
\end{lemma}
\begin{proof}
This follows directly from the definition of an earthquake and a valid earthquake. 
\end{proof}

Now we can show that switching the order of two earthquakes with certain validity conditions does not affect the final board state.
\begin{lemma}
\label{swappinglemma}
Consider a board $b$ and two earthquakes $E_1$ and $E_2$, affecting nodes $s_1$ and $s_2$ respectively. If $E_1, E_2$ is a valid sequence and $E_2$ is a valid earthquake when applied before $E_1$ then $E_2, E_1$ is also a valid sequence of earthquakes and 
\begin{equation*}
E_2 E_1 b = E_1 E_2 b
\end{equation*}
\end{lemma}
\begin{proof}
We will first show that $E_2, E_1$ is a valid sequence and then use this to show that the order in which the earthquakes occur does not matter.

We demonstrate this by contradiction. Assume  $E_2, E_1$ is an invalid sequence, that is $E_2 b(s_1)<4$. Since $E_1$ is a valid earthquake when applied directly to $b$ we know $b(s_1) \geq 4 $. 
Therefore we must have $b(s_1)<8$ and $s_2=s_1$. But then $E_1 b(s_1)<4$ and so $E_1, E_2$ would not be a valid sequence of earthquakes. This contradicts one our hypothesis and so $E_2, E_1$ is a valid sequence.

Now in order to demonstrate that the order in which the earthquakes are applied does not affect the final board state consider $E_2 E_1 b(x)$ for some $x \in \mathbb{Z} \times \mathbb{Z}$. Applying Lemma ~\ref{earthquakeredefine} we have 
\begin{align*}
E_2 E_1 b (x) = E_1(x)  + [[ x \in n(s_2) ]] - 4[[ x=s_2]] \\
= b(x) + [[ x \in n(s_1) ]] - 4[[x = s_1] + [[ x \in n(s_2) ]] - 4[[ x= s_2)]]  \\
= b(x) + [[ x \in n(s_2) ]] - 4[[x = s_2] + [[ x \in n(s_1) ]] - 4[[ x= s_1)]]\\
= E_2 b(x) + [[ x \in n(s_1) ]] - 4[[ x = s_1]] \\
= E_1 E_2 b(x) 
\end{align*}
The critical step in this proof is recognizing that because $E_1$ and $E_2$ are valid earthquakes regardless of their ordering the algebra above was valid.
\end{proof}

@@

\begin{definition}
A $d$-tree is a graph composed of a parent node connected by edges to $d-1$ unique children nodes.Those children each have $d-1$ unique children and so on.  
\end{definition}
\begin{figure}[!h]
\label{4tree}
\centering
\includegraphics[width=0.4\linewidth]{4tree}
\caption{A 4-tree}
\end{figure}
We note that every node other than the original parent node has $d$ neighbors. In order to simplify our process we would like the graph to be regular and so we make the following definition. 
\begin{definition}
A joined $d$-tree is two $d$-trees connected at their parent nodes by an edge.
\end{definition}
\begin{figure}[!h]
\label{joined4tree}
\centering
\includegraphics[width=0.8\linewidth]{joined4tree}
\caption{A joined 4-tree}
\end{figure}

@@

In order to keep our notation compact we once again define a function that outputs the neighbors of a node
\begin{definition} 
Let $n(i,j)d$ be the set of neighbors of node $(i,j)$ 
\begin{equation*}
n(i,j) = \{ (x_1, y_1); (x_1 = i+1 \text{ or } x_1=i-1), \text{ and }(x_1, y_1) \in G_d\}
\end{equation*}
\end{definition}

We now formalize the board of poker chips. Note that this definition only differs from our initial definition in that the input value is a coordinate of a node in a joined $d$-tree rather than a coordinate of the cartesian grid.
\begin{definition}
We define a $d$-board to be a map from the set of valid coordinates to the nonnegative integers, that is 
\begin{equation}
b:  G_d \rightarrow \mathbb{Z}_{\geq 0}
\end{equation}
We write $\mathcal{B}(d)$ to refer to the set of all $d$-boards.  
\end{definition}

Recalling that $n(s)$ refers to the neighbors of $s$ we define our earthquake operation in terms of $d$-boards. It functions analogously to the earthquake for the standard board, but instead of hitting nodes that have more than 4 chips it hits nodes that have more than $d$ chips. 
\begin{definition}
An earthquake takes as an input an $d$-board and a node and outputs an $d$-board so we write $E: \mathcal{B}(d) \times G_d \rightarrow \mathcal{B}(d)$. If $s$ is the node input and $b(s)>d$ then 
\begin{align*}
E_s b(s) = b(s)- d \\
E_s b(t) = b(t) +1 \quad \forall t \in n(s)
\end{align*}
Otherwise $b(s)<d$ and $E_s b=b$ as the earthquake does not move any chips.
\end{definition}

Now that we have formalized the earthquake operator we can introduce a precise definition of stability of a board. 
\begin{definition}
We say that a $d$-board is stable if its values are strictly less than $d$.
\end{definition}
This is equivalent to saying all earthquakes have no effect on the board. We also make a precise definition for a finite board.
\begin{definition} 
We say that a $d$-board is finite if the total number of chips on it is finite.
\end{definition}

@@

We once again define a spread function 
\begin{definition}
To define the spread function we once again define a distance on our graph.  The distance for a node $s$ 
$\Psi$ is a function from the finite $d$-boards and is defined
\begin{equation}
\Psi(b) = \sum_{(i,j) \in G_d} b(i,j) |i|^2
\end{equation}
Note that this is well defined for trees which contain a finite number of chips. 
\end{definition} 
We can once again think of this spread function as the sum of the number of chips on every node weighted by the taxi-cab distance from the node to the origin.
